---
title: "Model Selection for Temporal Extrapolation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Selection for Temporal Extrapolation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

When extrapolating treatment effects to future time periods using temporal models (Path 2 in the paper), the choice of functional form $f(g,t;\gamma)$ is critical. A linear-in-time model may underfit if effects are accelerating; a quadratic may overfit and produce unstable extrapolations. This vignette demonstrates the **time-series cross-validation framework** from Section 5.2 of the paper, which evaluates candidate models based on their extrapolation performance on held-out periods.

This vignette shows how to:

1. Specify candidate temporal models
2. Run time-series cross-validation to compute mean squared prediction error (MSPE)
3. Select the best model based on MSPE
4. Use the selected model for extrapolation
5. Optionally average across models with exponential weights

## Setup

```{r}
library(extrapolateATT)
set.seed(20260304)
```

## Example: Quadratic vs Linear Extrapolation

We'll simulate data where the true group-time effects follow a **quadratic trend**, then use cross-validation to see if the procedure correctly selects the quadratic model over linear.

### Step 1: Generate Data with Quadratic Trend

```{r}
# Parameters
n <- 200  # Sample size
n_groups <- 2
times <- 1:7
groups <- c(0, 1)

# Generate data with quadratic temporal pattern
# True DGP: tau_{gt} = 0.1 + 0.05*t + 0.02*t^2
data_list <- list()
phi_list <- list()
idx <- 1

for (g in groups) {
  for (t in times) {
    # Quadratic trend
    tau_true <- 0.1 + 0.05 * t + 0.02 * t^2

    # Add small noise
    tau_hat <- tau_true + rnorm(1, sd = 0.02)

    data_list[[idx]] <- data.frame(
      g = g,
      t = t,
      tau_hat = tau_hat,
      k = t - g
    )

    # Generate EIF vectors (influence functions)
    phi_list[[idx]] <- rnorm(n, mean = 0, sd = 0.15)
    idx <- idx + 1
  }
}

# Combine into gt_object structure
data <- do.call(rbind, data_list)

gt_object <- list(
  data = data,
  phi = phi_list,
  times = sort(unique(data$t)),
  groups = sort(unique(data$g)),
  event_times = sort(unique(data$k)),
  n = n
)
class(gt_object) <- c("gt_object", "extrapolateATT")

print(gt_object$data)
```

### Step 2: Build Candidate Models

We'll test three models: linear, quadratic, and a constant baseline.

```{r}
# Built-in models (linear and quadratic)
models <- build_model_specs(c("linear", "quadratic"))

# Optionally add a custom constant model
constant_model <- list(
  constant = list(
    h_fun = function(times, future_time) {
      function(tau_g) mean(tau_g)
    },
    dh_fun = function(times, future_time) {
      rep(1 / length(times), length(times))
    },
    name = "constant"
  )
)

models <- build_model_specs(c("linear", "quadratic"), custom_models = constant_model)

names(models)
```

### Step 3: Run Time-Series Cross-Validation

We hold out the last 1, 2, and 3 periods as test sets, fit each model on earlier data, and compute MSPE on the held-out periods.

```{r}
cv_result <- cv_extrapolate_ATT(
  gt_object,
  model_specs = models,
  horizons = 1:3,           # Test 1, 2, and 3 periods ahead
  future_value = 8,         # Actual future time we care about
  time_scale = "calendar",
  compute_coverage = FALSE  # Skip for speed (can enable for coverage metrics)
)

print(cv_result)
```

### Step 4: Examine Results

```{r}
# Detailed summary
summary(cv_result)

# Visualize MSPE by horizon
plot(cv_result)
```

The quadratic model should have the lowest MSPE since the true DGP is quadratic.

### Step 5: Select and Use Best Model

```{r}
# Select best model by MSPE
best_model_name <- select_best_model(cv_result, criterion = "mspe")
cat("Selected model:", best_model_name, "\n")

# Extract the best model specification
best_spec <- models[[best_model_name]]

# Use it to extrapolate to future time 8
result <- extrapolate_ATT(
  gt_object,
  h_fun = best_spec$h_fun,
  dh_fun = best_spec$dh_fun,
  future_value = 8,
  time_scale = "calendar",
  per_group = TRUE
)

print(result$tau_g_future)
```

### Compare with Other Models

To see how much the model choice matters, let's compare extrapolations from all three models:

```{r}
# Extrapolate with each model
extrapolations <- lapply(names(models), function(model_name) {
  spec <- models[[model_name]]
  res <- extrapolate_ATT(
    gt_object,
    h_fun = spec$h_fun,
    dh_fun = spec$dh_fun,
    future_value = 8,
    time_scale = "calendar",
    per_group = TRUE
  )

  data.frame(
    model = model_name,
    g = res$tau_g_future$g,
    tau_future = res$tau_g_future$tau_future
  )
})

comparison <- do.call(rbind, extrapolations)
print(comparison)

# True value at t=8 for comparison
true_tau_8 <- 0.1 + 0.05 * 8 + 0.02 * 8^2
cat("\nTrue tau at t=8:", round(true_tau_8, 3), "\n")
```

The quadratic model should be closest to the truth.

## Model Averaging

Instead of selecting a single model, we can average across all models using exponential weights based on MSPE. Models with lower MSPE get higher weight.

```{r}
# Model-averaged extrapolation
avg_result <- average_models(
  cv_result,
  gt_object,
  future_value = 8,
  time_scale = "calendar",
  temperature = 1,  # Controls weight concentration (default)
  per_group = TRUE
)

print(avg_result)

# Check the weights
cat("\nModel weights:\n")
print(avg_result$weights)
```

The model-averaged estimate combines predictions from all models, weighted by their cross-validated performance. This acknowledges model uncertainty and can improve robustness when multiple models perform similarly.

### Temperature Parameter

The `temperature` parameter controls how concentrated the weights are:

- Low temperature (e.g., 0.1): Winner-take-all (nearly all weight on best model)
- Default (1.0): Standard exponential weighting
- High temperature (e.g., 10): Nearly uniform weights (equal averaging)

```{r}
# Winner-take-all weighting
avg_concentrated <- average_models(
  cv_result, gt_object, future_value = 8,
  time_scale = "calendar", temperature = 0.1, per_group = TRUE
)

# Uniform weighting
avg_uniform <- average_models(
  cv_result, gt_object, future_value = 8,
  time_scale = "calendar", temperature = 10, per_group = TRUE
)

cat("Concentrated weights (temp=0.1):\n")
print(avg_concentrated$weights)

cat("\nUniform weights (temp=10):\n")
print(avg_uniform$weights)
```

## Coverage-Based Model Selection

If you care about uncertainty quantification as well as point prediction, you can select models based on both MSPE and coverage. This requires computing confidence intervals for test predictions (set `compute_coverage = TRUE`).

```{r, eval=FALSE}
# Run CV with coverage computation (slower)
cv_with_coverage <- cv_extrapolate_ATT(
  gt_object,
  model_specs = models,
  horizons = 1:2,
  future_value = 8,
  time_scale = "calendar",
  compute_coverage = TRUE,  # Enable coverage
  level = 0.95
)

summary(cv_with_coverage)

# Select model with good coverage among those with low MSPE
best_with_coverage <- select_best_model(
  cv_with_coverage,
  criterion = "coverage",
  tolerance = 0.15  # Consider models within 15% of best MSPE
)

cat("Best model (coverage criterion):", best_with_coverage, "\n")

# Or use combined criterion (ranks both MSPE and coverage)
best_combined <- select_best_model(
  cv_with_coverage,
  criterion = "combined"
)

cat("Best model (combined criterion):", best_combined, "\n")
```

## Practical Workflow Summary

1. **Specify candidate models:** Use `build_model_specs()` for common models (linear, quadratic), or add custom models.

2. **Run cross-validation:** Call `cv_extrapolate_ATT()` with a range of horizons (e.g., 1-3 periods). The function holds out each horizon as a test set and computes MSPE.

3. **Examine results:** Use `print()`, `summary()`, and `plot()` to understand which models perform best. Check if MSPE differences are substantial.

4. **Select or average:**
   - If one model clearly dominates, select it with `select_best_model()`.
   - If multiple models perform similarly, use `average_models()` to hedge against model uncertainty.

5. **Extrapolate:** Use the selected model or averaged result to predict future ATTs.

6. **Report sensitivity:** Include extrapolations from the top 2-3 models to show how sensitive results are to functional form choice.

## Recommendations

- **Start simple:** Test linear and quadratic models as a baseline.
- **Use multiple horizons:** Don't rely on a single horizon (e.g., h=1). Test 2-3 horizons to see if a model extrapolates well both near-term and longer-term.
- **Check coverage:** If inference is critical, enable coverage computation to ensure your selected model produces well-calibrated confidence intervals.
- **Model average when uncertain:** If MSPE differences are small (e.g., ratio < 1.5), model averaging is safer than picking a single model.
- **Plot diagnostics:** Always visualize MSPE by horizon and look at individual predictions (stored in `cv_result$predictions`).

## Limitations

Time-series CV evaluates extrapolation performance **within the observed time window**. It cannot test extrapolation to periods that differ fundamentally from observed periods (e.g., due to regime changes, policy reforms, or distributional shifts). See paper Section 5.2 for a discussion of when CV is valid and when structural assumptions are needed.

## References

This vignette implements the model selection framework from **Section 5.2** of:

> Agniel, D. (2026). Estimating policy effects in the presence of heterogeneity. *Manuscript in preparation*.

Key theoretical results:
- **Section 5.1:** EIF propagation through temporal models
- **Section 5.2:** Time-series cross-validation for model selection
- **Section 7.7:** Simulation study comparing linear, quadratic, and spline models
